---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
# Reproducible Research: Peer Assessment 1
# Craig Lewis
# June 15, 2015

## Loading and preprocessing the data
```{r ,echo=FALSE,results="hide",warning=FALSE,message=FALSE}
library(dplyr)
library(xtable)
step<-read.csv("activity.csv")
```



## What is mean total number of steps taken per day?
```{r showtable,results="asis"}
# summarize data & fixup column name
date_table<-summarise(group_by(step,date),sum(steps,na.rm=TRUE))
names(date_table)[2]<-"StepSum"
#output table by day
dt<-xtable(date_table)
print ("Number of steps taken each day:")
print(dt,type="html")

#print histogram
hist(date_table$StepSum,breaks=30,xlab="Steps",main="Histogram of Steps")


#calculate mean & median
daily_mean <- round(mean(date_table$StepSum),2)
daily_median<-median(date_table$StepSum)

```

The mean total number of steps taken per day is `r daily_mean` steps. The median number of steps is `r daily_median` steps.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
